Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-02-06 paper.tex
merge - not sure how I messed this up :/

\newcommand{\system}{\textsc{HelmholtzHacker}~}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2018}

\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\usepackage{listings}
\usepackage{amsthm}
% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfig} 
\usepackage{fancyvrb}

@@

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Figure~\ref{graphicalModel}(a), finding the $\mathcal{D}^*,\theta^*$ solving:
\begin{align}\label{intractableObjectives}
  \nonumber\probability[X|\mathcal{D},\theta] &= \prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \nonumber\probability[\mathcal{D}|X]&\propto \probability[\mathcal{D}] \int \mathrm{d}\theta P(\theta|\mathcal{D})\probability[X|\mathcal{D},\theta]\\
  \mathcal{D}^* &= \argmax_{\mathcal{D}}\probability[\mathcal{D}|X]\\
\nonumber  \theta^*& =\argmax_\theta P(\theta |\mathcal{D}^*)\probability[X|\mathcal{D}^*,\theta]
\end{align}
The above equations summarize the problem from the point of view of an ideal Bayesian learner.
However, Eq.~\ref{intractableObjectives}
is wildly intractable because evaluating $\probability[X|\mathcal{D},\theta]$ involves
summing over the  infinite set of all possible programs.
In practice we will only ever be able to sum over a finite set of programs.
So, for each task, we define a finite set of programs, called a \emph{frontier}, and only marginalize over the frontiers:
\\\noindent\textbf{Definition.} A \emph{frontier of task $x$}, written $\mathcal{F}_x$,
is a finite set of programs where $\probability[x|p] > 0$ for all $p\in \mathcal{F}_x$.

With the frontiers in hand, we can define the following intuitive lower bound on the likelihood of the tasks, which we call $\mathcal{L}$:
\begin{align}
  \probability[X|\mathcal{D},\theta]\geq J\triangleq\prod_{x\in X} \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{align}
along with a corresponding lower bound on the marginal likelihood of a DSL, which we call $$
\begin{align}
  \probability[\mathcal{D}|X]\geq 
  \end{align}

If we had a $(\mathcal{D},\theta)$ solving Eq.~\ref{intractableObjectives}, then we could recover the most likely program for task $x$ by maximizing $\probability[x|p] \probability[p|\mathcal{D},\theta]$.
Through this lens we now take as our goal to solve Eq.~\ref{intractableObjectives}.
But even \emph{evaluating} Eq.~\ref{intractableObjectives} is intractable because it involves summing over the infinite set of all possible programs, as an ideal Bayesian learner would.
In practice, we must instead marginalize over
some finite set of programs.

In general, programs are hard-won: finding even a single program that explains a given observation presents a daunting combinatorial search problem.

Now in theory we would like to sum over the entire infinite space
of all programs -- but this is of course impossible.

%% In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
%% Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
%% \begin{align}
%%   A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
%% &+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
%%   \end{align}

With this fact in mind, we will instead maximize the following tractable lower bound on Eq.~\ref{AIC},
which we call $J$:
\begin{equation}
J = \log \probability[\mathcal{D},\theta] + \sum_{x\in X}\log \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]\label{lowerBound}
\end{equation}
This lower bound depends on sets of programs, $\left\{\mathcal{F}_x \right\}_{x\in X}$:

We maximize $J$ by alternating maximization w.r.t. the frontiers and the DSL:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.

@@

\textbf{Neural recognition model: tractably maxing $J$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large.  With $q$ in hand we can find programs for
the frontier of a task $\mathcal{F}_x$ by searching for programs that maximize
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta^{(x)}$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
This approach implements an amortized
inference scheme~\cite{le2016inference,ritchie2016deep} for the generative model in
Figure~\ref{graphicalModel}.

@@

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
We represent programs as $\lambda$-calculus expressions.
$\lambda$-calculus is a formalism for expressing functional programs
that closely resembles the Lisp programming language.
$\lambda$-calculus includes variables, function application, and the ability to create new functions.
Throughout this paper we will write $\lambda$-calculus expressions in Lisp syntax.
Our programs are all strongly typed.
We use the Hindley-Milner polymorphic typing system~\cite{pierce} which is used in functional languages like OCaml.
Variables that range over types are always written using lowercase Greek letters
and we write $\alpha\to \beta$ to mean a function that takes as input type $\alpha$
and returns something of type $\beta$.
We use the notation $p:\tau$ to mean that the $\lambda$-calculus expression $p$
has the type $\tau$.
For example, to describe the type of the identity function
we would say \texttt{(lambda (x) x)}$:\alpha\to \alpha$.
We say a type $\alpha$ \emph{unifies} with $\tau$ if every expression
$p:\alpha$ also satisfies $p:\tau$. Furthermore, the act of \emph{unifying}
a type $\alpha$ with $\tau$ is to introduce constraints on the type
variables of $\alpha$ to ensure that $\alpha$ unifies with $\tau$.
See Section~\ref{programrepresentation} for more detail on program representation.
With this notation in hand we can now define DSLs:

@@

Exact maximization of $J$ (Eq.\ref{lowerBound}) w.r.t. $(\mathcal{D},\theta)$
is intractable,
so we take a more heuristic approach.
The strategy is to search locally through the space of DSLs,
proposing small local changes to $\mathcal{D}$ until $J$ fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting subexpressions from
programs already in the frontier.
These extracted subexpressions
are fragments of the original programs, and can introduce new variables (Figure~\ref{fragmentExample}),
which then become new functions in the DSL.
Our model of subexpressions is loosely related to Fragment
Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing},
but context-sensitive.

We define a prior distribution over DSLs which penalizes the sizes of the $\lambda$-calculus expressions in the DSL, and put a Dirichlet prior over the weight vector:
\begin{align*}
  \probability[\mathcal{D}]&\propto\exp\left(\lambda\sum_{p\in \mathcal{D}}\text{size}(p) \right)\\
  P(\theta|\mathcal{D})& = \text{Dir}(\theta|\alpha)
\end{align*}
where $\text{size}(p)$  measures the size of the syntax tree of program $p$,
$\lambda$ is a hyperparameter that acts as a regularizer on the size of the DSL,
and $\alpha$ is a concentration parameter controlling the smoothness of the prior over $\theta$.
Algorithm~\ref{grammarInductionAlgorithm} specifies the DSL induction algorithm.

>>>>

\newcommand{\system}{\textsc{HelmholtzHacker}~}
\newcommand{\lowerBound}{\mathscr{L}}
% Use the following line for the initial blind version submitted for review:
\usepackage{icml2018}

\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{mathrsfs}
\usepackage{listings}
\usepackage{amsthm}
% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfig} 
\usepackage{fancyvrb}

@@

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Figure~\ref{graphicalModel}(a). Writing $J$ for the joint probability of $(\mathcal{D},\theta)$, we want the $\mathcal{D}^*,\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)&\triangleq \probability[\mathcal{D}]P(\theta)\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* &= \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta \\
\nonumber  \theta^*& =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}
The above equations summarize the problem from the point of view of an ideal Bayesian learner.
However, Eq.~\ref{intractableObjectives}
is wildly intractable because evaluating $J(\mathcal{D},\theta)$ involves
summing over the  infinite set of all possible programs.
In practice we will only ever be able to sum over a finite set of programs.
So, for each task, we define a finite set of programs, called a \emph{frontier}, and only marginalize over the frontiers:
\\\noindent\textbf{Definition.} A \emph{frontier of task $x$}, written $\mathcal{F}_x$,
is a finite set of programs where $\probability[x|p] > 0$ for all $p\in \mathcal{F}_x$.

Using the frontiers we  define the following intuitive lower bound on the joint probability, which we call $\lowerBound$:
\begin{align}
 J\geq \lowerBound\triangleq\probability[\mathcal{D}]P(\theta)\prod_{x\in X} \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{align}

%% If we had a $(\mathcal{D},\theta)$ solving Eq.~\ref{intractableObjectives}, then we could recover the most likely program for task $x$ by maximizing $\probability[x|p] \probability[p|\mathcal{D},\theta]$.
%% Through this lens we now take as our goal to solve Eq.~\ref{intractableObjectives}.
%% But even \emph{evaluating} Eq.~\ref{intractableObjectives} is intractable because it involves summing over the infinite set of all possible programs, as an ideal Bayesian learner would.
%% In practice, we must instead marginalize over
%% some finite set of programs.

%% In general, programs are hard-won: finding even a single program that explains a given observation presents a daunting combinatorial search problem.

%% Now in theory we would like to sum over the entire infinite space
%% of all programs -- but this is of course impossible.

%% In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
%% Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
%% \begin{align}
%%   A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
%% &+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
%%   \end{align}
Mechanistically, \system does approximate MAP inference in the generative model of Figure~\ref{graphicalModel}(a) by maximizing this lower bound on the joint probability.
We maximize $\lowerBound$ by alternatingly maximizing w.r.t. the frontiers and the DSL:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $\lowerBound$ w.r.t. the frontiers.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the frontiers so that $\lowerBound$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $\int \lowerBound\;\mathrm{d}\theta$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\int \lowerBound\;\mathrm{d}\theta$.
Once we have a DSL $\mathcal{D}$ we can update $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{F}_x \right\})$.

@@

\textbf{Neural recognition model: tractably maxing $\lowerBound$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large.  With $q$ in hand we can find programs for
the frontier $\mathcal{F}_x$ by searching for programs that maximize
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta^{(x)}$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
%% This approach implements an amortized
%% inference scheme~\cite{le2016inference,ritchie2016deep} for the generative model in
%% Figure~\ref{graphicalModel}.

Our lower bound $\lowerBound$ is unconventional,
and one might wonder why we do not instead maximize an ELBO-style bound like in a VAE or in the EM algorithm.
For the interested reader, we explain in the supplement that maximizing an ELBO-style bound
leads to a pathological behavior that would cause the algorithm to easily become trapped in local optima.

@@

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
We represent programs as $\lambda$-calculus expressions.
$\lambda$-calculus is a formalism for expressing functional programs
that closely resembles the Lisp programming language.
$\lambda$-calculus includes variables, function application, and the ability to create new functions.
Throughout this paper we will write $\lambda$-calculus expressions in Lisp syntax.
Our programs are all strongly typed.
We use the Hindley-Milner polymorphic typing system~\cite{pierce} which is used in functional languages like OCaml.
Variables that range over types are always written using lowercase Greek letters
and we write $\alpha\to \beta$ to mean a function that takes as input type $\alpha$
and returns something of type $\beta$.
We use the notation $p:\tau$ to mean that the $\lambda$-calculus expression $p$
has the type $\tau$.
For example, to describe the type of the identity function
we would say \texttt{(lambda (x) x)}$:\alpha\to \alpha$.
We say a type $\alpha$ \emph{unifies} with $\tau$ if every expression
$p:\alpha$ also satisfies $p:\tau$. Furthermore, the act of \emph{unifying}
a type $\alpha$ with $\tau$ is to introduce constraints on the type
variables of $\alpha$ to ensure that $\alpha$ unifies with $\tau$.
See Supplement for more detail on program representation.
With this notation in hand we can now define DSLs:

@@

Exact maximization of $\lowerBound$ (Eq.\ref{lowerBound}) w.r.t. $(\mathcal{D},\theta)$
is intractable,
so we take a more heuristic approach.
The strategy is to search locally through the space of DSLs,
proposing small local changes to $\mathcal{D}$ until $\lowerBound$ fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting subexpressions from
programs already in the frontier.
These extracted subexpressions
are fragments of the original programs, and can introduce new variables (Figure~\ref{fragmentExample}),
which then become new functions in the DSL.

%% Our model of subexpressions is loosely related to Fragment
%% Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing},
%% but context-sensitive.

The idea of storing and reusing
fragments of expressions comes from Fragment Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing}.
Somehow I need to clarify that this is not like an adapter grammar, and that previous work has done adapter grammar like things... In fact this might be a baseline, because with the right parameter settings our algorithm can revert to an adapter grammar like phase.

We define a prior distribution over DSLs which penalizes the sizes of the $\lambda$-calculus expressions in the DSL, and put a Dirichlet prior over the weight vector:
\begin{align*}
  \probability[\mathcal{D}]&\propto\exp\left(-\lambda\sum_{p\in \mathcal{D}}\text{size}(p) \right)\\
  P(\theta|\mathcal{D})& = \text{Dir}(\theta|\alpha)
\end{align*}
where $\text{size}(p)$  measures the size of the syntax tree of program $p$,
$\lambda$ is a hyperparameter that acts as a regularizer on the size of the DSL,
and $\alpha$ is a concentration parameter controlling the smoothness of the prior over $\theta$.
Algorithm~\ref{grammarInductionAlgorithm} specifies the DSL induction algorithm.
